A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2014-11-21 journal.tex
Near a final Draft.

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}

@@

\begin{document}
%opening
 \title{Prototyping Collision-Free MAC Protocols in Real Hardware}

%A more simple output, useful when involving people from different affiliations
  %\author{
    %  \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}\\
      %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
      %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
  %}

\author{Luis Sanabria-Russo \\
		NeTS Research Group at\\
		Universitat Pompeu Fabra, Barcelona, Spain\\
		\texttt{Luis.Sanabria@upf.edu}}

%This is the style of three columns, as indicated in IEEEtran
% \author{\IEEEauthorblockN{Luis Sanabria-Russo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: luis.sanabria@upf.edu}
%  \and
%  \IEEEauthorblockN{Jaume Barcelo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: cristina.cano@upf.edu} 
%  \and
%  \IEEEauthorblockN{Boris Bellalta}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: boris.bellalta@upf.edu}}

@@

@@

\boldmath Collisions are a main cause of throughput degradation in WLANs. The current contention mechanism used in this type of network called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) uses a Binary Exponential Backoff (BEB) mechanism to delay each contender attempt of transmitting, effectively reducing the collision probability. Nevertheless, CSMA/CA relies on a random backoff which in principle is unable to eliminate collisions, resulting in a network throughput degradation as more contenders attempt to share the channel. Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successfull transmissions. CSMA/ECA is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and provide short-term throughput fairness among contenders.

This work reviews CSMA/ECA mechanisms and provides insightful simulations and the first real-life tests results that reveal its benefits over CSMA/CA under different network traffic conditions.

@@

\begin{IEEEkeywords}
OpenFWWF, WMP, MAC, Collision-free, CSMA/ECA.
\end{IEEEkeywords}

\section{Introduction}\label{introduction}
\section{Related Work}\label{relatedWork}
\section{Simulation Results}\label{simulations}
\section{Prototyping CSMA/ECA}\label{prototype}
\section{CSMA/ECA in real hardware results}\label{prototypeResults}
\section{Conclusions}\label{conclusions}
\section{Acknowledgements}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
%Algorithms
\usepackage[linesnumbered,vlined]{algorithm2e}

\usepackage[hyphens]{url}
\usepackage{hyperref}

%underline text spanning several lines
\usepackage{soul}

%Put the references in a nice way so the empty space is not overly great.
\usepackage{flushend}

@@

%Footnote without a number
\makeatletter
\def\blfootnote{\xdef\@thefnmark{}\@footnotetext}
\makeatother

\usepackage{lipsum}

\begin{document}
%opening
 \title{A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios}

%A more simple output, useful when involving people from different affiliations
  \author{
      \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}}%\\
      %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
      %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
 % }

%\author{Luis Sanabria-Russo \\
%		NeTS Research Group at\\
%		Universitat Pompeu Fabra, Barcelona, Spain\\
%		\texttt{Luis.Sanabria@upf.edu}}

%This is the style of three columns, as indicated in IEEEtran
% \author{\IEEEauthorblockN{Luis Sanabria-Russo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: luis.sanabria@upf.edu}
%  \and
%  \IEEEauthorblockN{Jaume Barcelo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: jaume.barcelo@upf.edu} 
%  \and
%  \IEEEauthorblockN{Boris Bellalta}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: boris.bellalta@upf.edu}}

@@

\blfootnote{This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).\\
L. Sanabria-Russo (\href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}) , J. Barcelo (\href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B. Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.}
%(e-mails: \href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}, \href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}, \href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu})

@@

\boldmath Collisions are a main cause of throughput degradation in WLANs. The current contention mechanism used in IEEE 802.11 networks is called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA). It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability. Nevertheless, CSMA/CA relies on a random backoff that while effective and totally distributed, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel. Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner using a deterministic backoff after successful transmissions. Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule. CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness.

%Further, by applying simple extensions to the backoff mechanism of CSMA/ECA it is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and providing short-term throughput fairness among contenders.

This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes. Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed

%This work describes CSMA/ECA and its extensions that allow it to achieve a collision-free schedule with many contenders in non-saturated conditions, as well as its performance when coexisting with CSMA/CA nodes. Further, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.

@@

\begin{IEEEkeywords}
CSMA/ECA, WLAN, MAC, Collision-free, Clock Drift.
\end{IEEEkeywords}

\section{Introduction}\label{introduction}
Wireless Local Area Networks (WLANs or WiFi networks~\cite{802Standards}) are a popular solution for wireless connectivity, whether in public places, work environments or at home. This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), offering a good tradeoff between performance and costs, which is a main reason for its popularity.

The Medium Access Control (MAC) scheme used in WLANs is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) protocol. It has been widely adopted by manufacturers and consumers, making it inexpensive to implement and an ubiquitous technology. Nevertheless, the ever-growing throughput demands from upper layers are faced with a bottleneck at the WLANs' MAC~\cite{perahia2008ieee}, which by its nature is prone to collisions that degrade the overall performance as more nodes join the network.

The research community has pushed forward many alternatives to the current MAC in WLANs~\cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
barcelo2008lba,bellalta2009vtc,HE,CSMA_ECA,L_MAC2,hui2011epp,barcelo2011tcf}, but when a proposal deviates too much from CSMA/CA, or some time-critical operations are modified, its hardware implementation as part of WLANs' MAC often becomes unlikely~\cite{WMP}, with the standardisation process taking many years without certainty of approval~\cite{perahia2008ieee}.

A CSMA/CA replacement should be able to provide advantages in terms of throughput, spectrum efficiency and number of supported contenders. All of the aforementioned without sacrificing short-term throughput fairness. Furthermore, it must also serve existing users, which means it has to be backwards compatible.

%handle many contenders without sacrificing short-term throughput fairness.

A suitable candidate, and the one to be evaluated in this work, is called Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)~\cite{barcelo2008lba}. It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism. In CSMA/ECA, nodes use a deterministic backoff after successful transmissions, constructing a collision-free schedule among successful contenders in a fully decentralised way. This backoff mechanism ensures that more channel time is spent on successful transmissions rather than recovering from collisions, thus increasing the throughput of the network. Further enhancements (or extensions), like \emph{Hysteresis} and \emph{Fair Share}~\cite{research2standards} allow CSMA/ECA to support many more contenders in a collision-free schedule. Moreover, CSMA/ECA and its extensions are designed for allowing nodes to transmit as frequently as possible while keeping an even distribution of the available bandwidth among users.

The 802.11 High Efficiency WLAN (HEW) Study Group (SG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios}, specifically those usually encountered at stadiums or conference rooms. Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount. Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement. Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users. Furthermore, the impact of imperfect clocks over the deterministic backoff mechanism is also lacking.

This paper fills those gaps by extending~\cite{research2standards}, and consolidates the push of CSMA/ECA as a potential replacement of CSMA/CA for next generation WLANs. In detail, this paper provides the following contributions:

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
%	\item Overview of real hardware implementations of CSMA/ECA prototypes.
\end{itemize}

Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules. Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than the observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.

An overview of similar distributed and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}. CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}. Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results. An overview of CSMA/ECA real-hardware implementations is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}. Conclusions are drawn in Section~\ref{conclusions}.

\section{Related Work}\label{relatedWork}
Time in WLANs is divided into tiny empty slots of fixed length $\sigma_{e}$, collisions and successful slots of length $\sigma_{c}$ and $\sigma_{s}$, respectively. Collision and successful slots contain collisions or successful transmissions, making them several orders of magnitude larger than empty slots ($\sigma_{e}\ll\min(\sigma_{s},\sigma_{c}))$. One of the effects of collisions is the degradation of the network performance by wasting channel time on collisions slots.

Recent advances in the WLANs PHY~\cite{perahia2008ieee,6191306} push the research community towards the development of MAC protocols able to take advantage of a much faster PHY. By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput. Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a totally distributed fashion in order to avoid injecting extra control traffic that may reduce the data throughput.

The followings are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing a collision-free schedule.

\subsection{Zero Collision MAC}

Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a totally distributed way. It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $N$-slots in length. Backlogged stations pick a slot in the virtual cycle to attempt transmission. If two or more stations picked the same slot in the cycle, their transmissions will eventually collide. This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided. When all $M$ stations reserve a different slot, a collision-free schedule is achieved.

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $M>N$. Furthermore, if $N$ is overestimated ($N\gg M$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

\subsection{Learning-MAC}

Learning-MAC~\cite{L_MAC} is another MAC protocol able to build a collision-free schedule for many contenders. It does so defining a \emph{learning strength} parameter, $\beta\in(0,1)$. Each contender starts by picking a slot for transmission $s$ of the schedule $n$ of length $C$ at random with uniform probability. After a contender picks slot $s(n)$, its selection in the next schedule, $s(n+1)$, will be conditioned by the result of the current attempt. (\ref{success}) and (\ref{collisions-eq}) extracted from~\cite{L_MAC} show the probability of selecting the same slot $s(n)$ in cycle $n+1$.

\begin{equation} \label{success}
		\left. \begin{aligned}
			p_{s(n)}(n+1)&=1,\\
			p_{j}(n+1)&=0,
		\end{aligned}
	\right\}
	\qquad \text{\emph{Success}}
\end{equation}
\begin{equation} \label{collisions-eq}
	\left. \begin{aligned}
			p_{s(n)}(n+1)&=\beta p_{s(n)}(n),\\
			p_{j}(n+1)&=\beta p_{j}(n)+\frac{1-\beta}{C -1},
		\end{aligned}
	\right\}
	\qquad \text{\emph{Collision}}
\end{equation}
\\
for all $j\neq s(n),~j\in \{1,\dots ,C\}$. That is, if a station successfully transmitted in $s(n)$, it will pick the same slot on the next schedule with probability one. Otherwise, it follows~(\ref{collisions-eq}).

The selection of $\beta$ implies a compromise between fairness and convergence speed, which the authors determined $\beta=0.95$ to provide satisfactory results.

L-MAC is able to achieve higher throughput than the current MAC with a very fast convergence speed. Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to the current MAC or may require a centralised entity~\cite{barcelo2011tcf}.

Further extensions to L-MAC introduced an \emph{Adaptative} schedule length in order to increase the number of supported contenders in a collision-free schedule. This adaptive schedule length is doubled or halved depending on the presence of collisions or many empty slots per schedule, respectively. The effects of reducing the schedule length may provoke a re-convergence phase which can result in short-term fairness issues. Moreover, L-MAC nodes also require common knowledge of the start/end of the schedule.

\section{Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)}\label{introProtocol}

CSMA/ECA~\cite{barcelo2008lba} is a totally distributed and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{d}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.

Collisions are handled as in CSMA/CA, which is described in Algorithm~\ref{alg:CSMA_CA}. In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, $m$ is the maximum backoff stage of typical value $m=5$, and $R$ is the retransmissions limit with typical value $R=m+1$), then generates a random backoff, $B$. When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed. Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and using a random backoff, $B\in[0,2^{k}CW_{\min}]$. This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.

Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}). Figure~\ref{fig:BECA-example} shows an example of CSMA/ECA dynamics with four contenders.

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$; $k \leftarrow 0$\;
  %\tcc{Initialize backoff counter.}
  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$B>0$}{
        wait 1 slot\;
        $B \leftarrow B-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}{\label{collision}
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $k \leftarrow \min (k+1,m)$\;
        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;\label{finalCollision}
      }
    }
    $r \leftarrow 0$\;
    \colorbox{yellow}{$k \leftarrow 0$;}\\
	 \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$B \leftarrow \mathcal{U}[0,2^{k}{\rm{CW}_{min}}-1]$;}\\\label{randomBackoff}
    }
    {
      Discard packet\;
      $B \leftarrow \mathcal{U}[0,2^k {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\caption{\small{CSMA/CA. $r$ indicates the number of retransmission attempts, while $R$ is the maximum retransmission attempts limit. When it is reached, the packet waiting for transmission is dropped.}}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $k \leftarrow 0$\;
  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$B>0$}{
        wait 1 slot\;
        $B \leftarrow B-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $k \leftarrow \min (k+1,m)$\;
        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;
      }
    }
    $r \leftarrow 0$\;
    \colorbox{yellow}{$k \leftarrow 0$;}\\
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}{\rm{CW}_{min}})/2-1$\;}\label{deterministicBackoff}\\
	 $B \leftarrow B_{d}$\;
    }
    {
      Discard packet\;
      $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\caption{\small{CSMA/ECA.}}
\label{alg:CSMA_ECA}
\end{algorithm}

%where $k$ is reset ($k=0$) after each successful transmission and $m$ is the maximum backoff stage of typical value $m=5$

\begin{figure*}[tb]
\centering
  \includegraphics[width=0.8\linewidth]{figures/basicECA.eps}
  \caption{An example of the temporal evolution of CSMA/ECA in saturation}
  \label{fig:BECA-example}
\end{figure*}

In Figure~\ref{fig:BECA-example}, the \emph{STA-\#} labels represent stations willing to transmit. The horizontal lines represent a time axis with each number indicating the amount of empty slots left for the backoff to expire. Stations willing to transmit begin the contention for the channel by waiting a random backoff, $B$. The first outline highlights the fact that stations STA-3 and STA-4 will eventually collide because they have selected the same $B$. After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{d}=7$ in this case. By doing so, all successful STAs will not collide among each other in future cycles.

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA} (redrawn from~\cite{research2standards}), shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.

\begin{figure}[tb]
\centering
  \includegraphics[width=0.7\linewidth, angle=-90]{figures/DCF-vs-ECA/DCF-vs-ECA2.eps}
  \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
  \label{fig:BECA}
\end{figure}

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{d}$ ($B_{d}=7$ in the case of the figure). Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.

The \emph{JFI for CSMA/ECA} and \emph{JFI for CSMA/CA} curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for both protocols. Both protocols have a JFI equal to one, which suggests that the available throughput is shared evenly among all stations.

	\subsection{Supporting many more contenders}\label{moreContenders}
	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{d}$. When $N > B_{d}$, collisions reappear.

	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{d}$, we introduce \emph{Hysteresis}. Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{d}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$. This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule. This idea of a schedule is significantly different from the virtual schedule required by the protocols described in Section~\ref{relatedWork}, that is, CSMA/ECA with Hysteresis does not require a previous knowledge of the number of contenders, the result of previous transmissions or the start/end of the schedule, easing its implementation in real hardware.

	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{d}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).

	This unfairness issue is solved by instructing nodes at backoff stage $k$ to transmit $2^{k}$ packets on each attempt, thus proportionally compensating those nodes at higher backoff stages. This additional extension to CSMA/ECA is called \emph{Fair Share}. CSMA/ECA with Hysteresis and Fair Share will be referred to as CSMA/ECA$_{\text{Hys+FS}}$ in order to distinguish it from what was described until this point.

	The idea of allowing the transmission of more packets to stations that transmit less often was initially proposed by Fang et al. in~\cite{L_MAC}. It was later adapted to CSMA/ECA$_{\text{Hys}}$ and named Fair Share in~\cite{research2standards}. Figure~\ref{fig:fairness} shows the JFI for CSMA/CA as well as for CSMA/ECA$_{\text{Hys+FS}}$.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined.eps}
		\caption{Fairness comparison with nodes under saturation (simulation parameters can be found in Section~\ref{simulations})}
		\label{fig:fairness}
	\end{figure}

	In Figure~\ref{fig:fairness}, the only curve deviating from JFI = 1 is \emph{CSMA/ECA with Hysteresis}, suggesting an uneven partition of the channel access time among contenders (which is fixed with Fair Share).

	\begin{figure*}[tb]
	\centering
		\includegraphics[width=0.8\linewidth]{figures/csma_eca_different_backoff_short.eps}
		\caption{An example of the temporal evolution of CSMA/ECA$_{\text{Hys+FS}}$ in saturation ($CW_{\min}=16$)}
		\label{fig:ECA+Hyst}
	\end{figure*}

	Algorithm~\ref{alg:fullECA} shows an implementation of CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{d}=CW(k)/2$. The same behavior is followed by STA 3.

	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown in Figure~\ref{fig:ECA+H+F-throughput} extracted from~\cite{research2standards}. In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission. This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.

	\begin{algorithm}[tb]
	\While{the device is on}
	{
	  $r \leftarrow 0$ ; $k \leftarrow 0$\;
	  $b \leftarrow \mathcal{U}[0,2^k\rm{CW}_{min}-1]$\;
	  \While{there is a packet to transmit}{
	    %\tcc{Initialize $a$.}
	    \Repeat{($r = R$) or (success)}{
	      %\tcc{First, backoff.}
	      \While{$B>0$}{
	        wait 1 slot\;
	        $B \leftarrow B-1$\;
	      }
	      \colorbox{yellow}{Attempt transmission of $2^k$ packets;}\\
	      \If{collision}{
	        %\tcc{Random backoff.}
	        $r \leftarrow r+1$\;
	        $k \leftarrow \min (k+1,m)$\;
	        $B \leftarrow \mathcal{U}[0, 2^k  \rm{CW}_{min} -1]$\;
	      }
	    }
	    $r \leftarrow 0$\;
	    %$s \leftarrow 0$\;
	    \eIf{success}{
	      %\tcc{Random backoff.}
	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\\
		$B \leftarrow B_{d}$;\\
	    }
	    {
	      Discard packet\;
	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
	    }
	  }
	  Wait until there is a packet to transmit\;
	}
	\vspace{0.2cm}
	\caption{CSMA/ECA$_{\text{Hys+FS}}$}
	\label{alg:fullECA}
	\end{algorithm}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined.eps}
		\caption{Throughput comparison~\cite{research2standards} (simulation parameters can be found in Section~\ref{simulations})}
		\label{fig:ECA+H+F-throughput}
	\end{figure}

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.

	In scenarios where short-term fairness and the time between consecutive transmissions are not relevant, Fair Share can be replaced by \emph{Maximum Aggregation} (MaxAg), which will significantly improve the system throughput. In Maximum Aggregation all nodes aggregate as many packets as possible at every transmission, i.e., they send $2^m$ packets in each attempt.

	%greater throughput than CSMA/CA. Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA with nodes sending $2^m$ packets in each attempt. This behavior will be called CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+FS}}$ (Max Ag.) for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$, respectively from here forth.

	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$. Although CSMA/CA$_{\text{MaxAg}}$ performs maximum aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission. On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg.eps}
		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
		\label{fig:ECA-vs-DCF-maxAgg}
	\end{figure}

	To summarise the effects of using aggregation:
	\begin{itemize}
		\item It increases the aggregated throughput: because nodes are able to send multiple packet in each attempt, the system throughput is increased. Moreover, Fair Share compensates those nodes at higher backoff stages to ensure throughput fairness.
		\item Maximum aggregation supposes the deactivation of the Fair Share mechanism: performing maximum aggregation upon each transmission attempt is equivalent to having different schedule lengths and not compensating nodes at higher backoff stages. Although the aggregated throughput increases, this results in an uneven distribution of the channel time among contenders, which renders it unfair.
		\item Longer periods between transmission attempts: given that each transmission takes longer, the time between transmission attempts also increases. This may specially affect delay-sensitive applications.
	\end{itemize}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined.eps}
		\caption{Fairness comparision when using maximum aggregation}
		\label{fig:fairness-with-maximum-aggregation}
	\end{figure}

	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol.
